---
title: "Weighted Sliced Inverse Regression (WSIR): supervised dimension reduction for spatial transcriptomics data"
author:
  - Max Woollard
  - Linh Nghiem
  - Shila Ghazanfar
  - Pratibha Panwar
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: WSIR
vignette: |
  %\VignetteIndexEntry{WSIR supervised dimension reduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE
)
```

```{r setup, message=FALSE, warning=FALSE}
library(wSIR)
library(tidyverse)
library(vctrs)
library(umap)
library(scran)

library(MouseGastrulationData) # to load the data for this vignette
```

```{r}
set.seed(2024)
```

# Introduction

Weighted Sliced Inverse Regression (WSIR) is a supervised dimension reduction technique for spatial transcriptomics data. It creates a low-dimensional representation of the gene expression data using its spatial coordinates to obtain information about spatial patterns present in the gene expression data. It can be used to project new single-cell gene expression data into a low-dimensional space in such a way as to preserve the ability to predict a cell's spatial location from its low-dimensional embedding. 

## Vignette Goals

In this vignette, we will demonstrate how to use WSIR to obtain a low-dimensional embedding of gene expression data. We will then explore this embedding using the in-built functions in the package, but this low-dimensional matrix is more importantly used for your own downstream tasks that would benefit from a lower-dimensional representation of gene expression data that preserves information about each cell's spatial location. 

# Load Data

Load data from https://www.nature.com/articles/s41587-023-01766-z and prepare. 

```{r}
seqfish_data_sample1 <- LohoffSeqFISHData(samples = 1)
seqfish_data_sample1 = logNormCounts(seqfish_data_sample1) # log transform variance stabilising 
rownames(seqfish_data_sample1) <- rowData(seqfish_data_sample1)[,"SYMBOL"] # change rownames to gene symbols that are consistent across Google 
sample1_exprs = t(assay(seqfish_data_sample1, "logcounts")) # extract matrix of gene expressions 
sample1_coords = spatialCoords(seqfish_data_sample1)[,1:2]
colnames(sample1_coords) = c("x", "y")

seqfish_data_sample2 <- LohoffSeqFISHData(samples = 2)
seqfish_data_sample2 = logNormCounts(seqfish_data_sample2) # log transform variance stabilising 
rownames(seqfish_data_sample2) <- rowData(seqfish_data_sample2)[,"SYMBOL"] # change rownames to gene symbols that are consistent across Google 
sample2_exprs = t(assay(seqfish_data_sample2, "logcounts")) # extract matrix of gene expressions 
colnames(sample2_coords) = c("x", "y")
```

# Supervised dimension reduction with WSIR

Tune parameters to find optimal values for parameters slices, alpha

```{r}
plot = explore_wsir_params(exprs = sample1_exprs, coords = sample1_coords, alpha_vals = c(0,2,4,8), slice_vals = c(3,6,10))
plot
```

Perform WSIR

```{r}
wsir_obj = wSIR(exprs = sample1_exprs, coords = sample1_coords, slices = 6, alpha = 4) # create wsir object using optimal values for slices, alpha
```

Observe the genes that contribute the most to WSIR1 (first direction of WSIR) using top_genes function

```{r}
top_genes_obj = top_genes(WSIR = wsir_obj, highest = 8) # create top genes object
top_genes_plot = top_genes_obj$plot # select plot
top_genes_plot # print plot
```

Visualise the WSIR directions using visualise_wsir function

```{r}
vis_obj = visualise_wsir(coords = sample1_coords, WSIR = wsir_obj, dirs = 8) # create visualisations
vis_obj
```

Visualise UMAP plots of low-dimensional representation of sample1_exprs after WSIR, colouring points by the 4 genes that contribute the most to WSIR1 using vis_umap function

```{r}
umap_plots = vis_umap(exprs = sample1_exprs, WSIR = wsir_obj, highest_genes = top_genes_obj, n_genes = 4)
umap_plots
```

Projecting sample2_exprs into low-dimensional space using project_wSIR function. This low-dimensional embedding retains the ability to predict a cell's spatial location that was already present in the gene expression data.

```{r}
sample2_low_dim_exprs = project_wSIR(wsir = wsir_obj, newdata = sample2_exprs)
```
